---
title: "Project 1"
author: "Dhairav Chhatbar"
date: "6/3/2020"
output: html_document
---
```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(kableExtra)
library(scales)
```

This simple system recommends how a specific user will rate a movie given the user's past rating and how the overall movie has been rated

### Load Data
Load the the ratings and movie title datasets
```{r}
ratings_import <- read.csv("https://raw.githubusercontent.com/dhairavc/DATA612-RecommenderSystems/master/Project1/ratings.csv")[,1:3]
movie_titles <- read.csv("https://raw.githubusercontent.com/dhairavc/DATA612-RecommenderSystems/master/Project1/movies.csv")[,1:2]

```


### Crease Dense Subset
For educational purposes create a smaller dense dataset of 5 randomly selected movies with the most reviews and 12 random users with a max 2 movies not rated per user
```{r}
# Create a subset of random 5 movies which have grather than 150 reviews
m_filter <- ratings_import %>% select(movieId) %>% group_by(movieId) %>% summarise("RatingCount" = n()) %>% arrange(desc(RatingCount)) %>% filter(RatingCount > 150)
m_filter <- m_filter[sample(nrow(m_filter), 5),]

# Create subset of 12 random users which have rated atleast 3 of the 5 movies selected 
u_filter <- ratings_import %>% filter(movieId %in% m_filter$movieId) %>% select(userId) %>% group_by(userId) %>% summarise("Rated" = n()) %>% arrange(desc(Rated)) %>% filter(Rated %in% c(3,4,5))
rating_sample <- ratings_import %>% filter(movieId %in% m_filter$movieId) %>% filter(userId %in% sample(u_filter$userId,12))

# Join movie titles and remove movie Ids
rating_sample <- left_join(rating_sample, movie_titles, by = "movieId")
rating_sample <- rating_sample %>% select(-movieId) 


ratings <- spread(rating_sample, title, rating)
row.names(ratings) <- c(ratings$userId)
ratings <- ratings[,2:6]
ratings %>% kable() %>% kable_styling(bootstrap_options = "striped", full_width = F)
```
### Split Datasets
Split the existing dataset into training and testing datasets
```{r}

n <- nrow(ratings)
p_train <- .6
rowsintrain <- floor(n*p_train)
randomize_order <- ratings[sample(n),]
train_ratings <- randomize_order[1:rowsintrain,]
test_ratings <- randomize_order[(rowsintrain+1):n,]

#training dataset
train_ratings %>% kable() %>% kable_styling(bootstrap_options = "striped", full_width = F)

#testing dataset
test_ratings %>% kable() %>% kable_styling(bootstrap_options = "striped", full_width = F)

```
### Raw Average
Compute the raw average mean
```{r}
raw_average <- mean(as.matrix(train_ratings), na.rm = TRUE)
raw_average
```



### RMSE Function
Create function to calculate the Root Mean Squared Error (RMSE)
```{r}

RMSE_calc <- function(df, q) {
  sqrt(mean(as.matrix(((df-q)^2)),na.rm = TRUE))
}

```

### RMSE Calculation
Calculate the RMSE RMSE for the raw average for both testing and training datasets
```{r}

train_RMSE <- RMSE_calc(train_ratings, raw_average)
test_RMSE <- RMSE_calc(test_ratings, raw_average)

train_RMSE
test_RMSE
```



### Baseline Prediction for Training Dataset
Calculate user and movie bias and calculate the baseline predictions for the training dataset
```{r}
user_bias <- rowMeans(train_ratings, na.rm = TRUE) - raw_average
movie_bias <- colMeans(train_ratings, na.rm = TRUE) - raw_average

train_ratings_pred <- data.frame(matrix(rep(raw_average, nrow(train_ratings)*ncol(train_ratings)), nrow = nrow(train_ratings)))

colnames(train_ratings_pred) <- colnames(train_ratings)
rownames(train_ratings_pred) <- rownames(train_ratings)
train_ratings_pred <-round(sweep(train_ratings_pred,MARGIN = 2, STATS = movie_bias, FUN = "+")+user_bias,1)

#clip [.5, 5]
train_ratings_pred[which(train_ratings_pred < .5, arr.ind = TRUE)] <- .5
train_ratings_pred[which(train_ratings_pred > 5, arr.ind = TRUE)] <- 5

train_ratings_pred %>% kable() %>% kable_styling(bootstrap_options = "striped", full_width = F)

```


### Baseline Prediction for Test Dataset
Calculate user bias and calculate the baseline predictions for the training dataset using the movie bias from the training dataset
```{r}
user_bias_test <- rowMeans(test_ratings, na.rm = TRUE)-raw_average
test_ratings_pred <- data.frame(matrix(rep(raw_average, nrow(test_ratings)*ncol(test_ratings)), nrow = nrow(test_ratings)))
colnames(test_ratings_pred) <- colnames(test_ratings)
rownames(test_ratings_pred) <- rownames(test_ratings)
test_ratings_pred <-round(sweep(test_ratings_pred,MARGIN = 2, STATS = movie_bias, FUN = "+")+user_bias_test,1)

#clip [.5, 5]
test_ratings_pred[which(test_ratings_pred < .5, arr.ind = TRUE)] <- .5
test_ratings_pred[which(test_ratings_pred > 5, arr.ind = TRUE)] <- 5

test_ratings_pred %>% kable() %>% kable_styling(bootstrap_options = "striped", full_width = F)

```


### RMSE Calculation for Baseline Predictor
Calculate the RMSE for both predicted datasets
```{r}

train_RMSE_baseline <- RMSE_calc(train_ratings, train_ratings_pred)
test_RMSE_baseline <- RMSE_calc(test_ratings, test_ratings_pred)

train_RMSE_baseline
test_RMSE_baseline
```

### Conclusion
Summarize findings and calculate improvement over RMSE base just on the raw average
```{r}
raw_avg <- c(raw_average, raw_average)
RMSE <- c(train_RMSE, test_RMSE)
RMSE_b <- c(train_RMSE_baseline, test_RMSE_baseline)
imp <- percent(round(1-(RMSE_b/RMSE),5))

sum_df <- data.frame(raw_avg, RMSE, RMSE_b, imp)
colnames(sum_df) <- c("Raw Average", "RMSE", "RMSE Baseline", "Improvement")
rownames(sum_df) <- c("Train Ratings", "Test Ratings")

sum_df %>% kable() %>% kable_styling(bootstrap_options = "striped", full_width = F)

```

