---
title: "Project 4 - Accuracy and Beyond"
author: "Dhairav Chhatbar"
date: "6/27/2020"
output: html_document
---
The goal of this assignment is give you practice working with accuracy and other recommender system 
metrics
  
* As in your previous assignments, compare the accuracy of at least two recommender system 
algorithms against your offline data

* Implement support for at least one business or user experience goal such as increased 
serendipity, novelty, or diversity

* Compare and report on any change in accuracy before and after you’ve made the change in #2

* As part of your textual conclusion, discuss one or more additional experiments that could be 
performed and/or metrics that could be evaluated only if online evaluation was possible.  Also, 
briefly propose how you would design a reasonable online evaluation environment

### Load Data & Goals
In this assignment we will attempt the best recommender model on the Jester dataset which is a a rating of a set of jokes
```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(recommenderlab)
library(dplyr)
library(kableExtra)
data("Jester5k")
```

### Explore Data
Exploratory visuals show that this dataset has:
  
* 5000 users and 100 jokes
* Ratings are between -10 and +10, with a rating distribution leaning towards positive ratings
* The ratings are not normally distributed, therefore normalization of the ratings will be beneficial
* Most users have rated all 100 jokes making this dataset fairly dense with about 72% of the jokes rated across all 5000 users


```{r}
dim(Jester5k)

# Ratings
all_ratings <- getRatings(Jester5k)
ggplot() + geom_histogram(aes(all_ratings), binwidth = 1, col="white") + labs(title = "User Ratings", x="Ratings", y="Count") 

all_ratings_norm <- getRatings(normalize(Jester5k, "Center"))
ggplot() + geom_histogram(aes(all_ratings_norm), binwidth = 1, col="white") + labs(title = "User Ratings Normalized (Center)", x="Ratings", y="Count") 

summary(all_ratings_norm)

#Users
users <- rowCounts(Jester5k)
ggplot() + geom_histogram(aes(users), binwidth = 5, col="white") + labs(title = "Jokes Rated by Users", x="Jokes Rated", y="Number of Users")


#Density 
sum(users)/(dim(Jester5k)[1]*dim(Jester5k)[2])

```

#### Need to Remove
```{r}
Jester_Norm <-normalize(Jester5k, "Center")


eval_jokes <- evaluationScheme(data=Jester_Norm, method="split", train=.80, 
                               given=min(rowCounts(Jester_Norm))-3, goodRating = 5)
Jokes_rec_IBCF <- Recommender(data = getData(eval_jokes, "train"), method = "IBCF")
Jokes_pred_IBCF <- predict(Jokes_rec_IBCF, newdata = getData(eval_jokes, "known"), type="ratings", n=5)

calcPredictionAccuracy(Jokes_pred_IBCF, data = getData(eval_jokes, "unknown"))


head(getConfusionMatrix(evaluate(eval_jokes, method = "IBCF", n = seq(10, 100, 10))))
```

#### Build Models
We will simulate recommendations on 6 models all Center normalized:
  
* Item Based Content Filtering, w\ Cosine difference
* Item Based Content Filtering, w\ Pearson difference
* User Based Content Filtering, w\ Cosine difference
* User Based Content Filtering, w\ Pearson difference
* Singular Value Decomposition, k = 5
* Random recommendations to simulate baseline 

```{r, warning=FALSE}

eval_jokes <- evaluationScheme(data=Jester5k, method="split", train=.80, 
                               given=min(rowCounts(Jester_Norm))-3, goodRating = 5)
num_jokes_to_recommend <- 100

recc_algos <- list(
  IBCF_C = list(name = "IBCF", param = list(method = "Cosine", normalize = "Center")),
  IBCF_P = list(name = "IBCF", param = list(method = "Pearson", normalize = "Center")),
  UBCF_C = list(name = "UBCF", param = list(method = "Cosine", normalize = "Center")),
  UBCF_P = list(name = "UBCF", param = list(method = "Pearson", normalize = "Center")),
  m_SVD = list(name = "SVD", param = list( k=5, normalize = "Center")),
  random = list(name = "RANDOM", param=NULL)
)

recommendation_results <- evaluate(eval_jokes, method = recc_algos, n = c(1,5, seq(10, num_jokes_to_recommend, 10)))


```


#### Acucracy and ROC


```{r}

plot(recommendation_results, annotate = TRUE, legend = "topleft") 
title("ROC curve")


plot(recommendation_results, "prec/rec", annotate = TRUE, legend = "bottomright") 
title("Precision-recall")


```
### Parameter Optimization

```{r}
vector_k <- c(5, 10, 20, 30, 40)
models_to_evaluate <- lapply(vector_k, function(k){
  list(name = "SVD", param = list(normalize = "Center", k = k)) 
}) 
names(models_to_evaluate) <- paste0("IBCF_k_", vector_k)

n_recommendations <- c(1, 5, seq(10, 100, 10))
list_results <- evaluate(x = eval_jokes, method = models_to_evaluate, n = n_recommendations)


plot(list_results, annotate = 1, legend = "topleft") 
title("ROC curve")
plot(list_results, "prec/rec", annotate = 1, legend = "bottomright")
title("Precision-recall")

```

